---
output:
  html_document: default
  pdf_document: default
---
Red Wine Exploration by Hassan Jummah
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(dplyr)
library(psych)
library(pander)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data

df <- read.csv("wineQualityReds.csv")
df <- select (df,-c(X))

bestWineDf <- subset(df, quality == max(quality))
worstWineDf <- subset(df, quality == min(quality))

```

## This dataset is public available for research. The details are described in [Cortez et al., 2009]
  Please include this citation if you plan to use this database:

  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
                [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib

#### In the above reference, two datasets were created, using red and white wine samples. The inputs include objective tests (e.g. PH values) and the output is based on sensory data (median of at least 3 evaluations made by wine experts). Each expert graded the wine quality between 0 (very bad) and 10 (very excellent). Several data mining methods were applied to model these datasets under a regression approach. The support vector machine model achieved the best results. Several metrics were computed: MAD, confusion matrix for a fixed error tolerance (T), etc. Also, we plot the relative importances of the input variables (as measured by a sensitivity analysis procedure).
  


# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
cat(" Number of Rows = ", nrow(df), "\n Number of Columns = ", ncol(df))
```

```{r echo=FALSE}
# Drop the columns of the dataframe

#Transforming Quality from an Integer to a Factor
df$quality <- factor(df$quality, ordered = T)
#Creating a new Factored Variable called 'Rating'
df$rating <- ifelse(df$quality < 5, 'Below average', ifelse(
  df$quality < 7, 'Average', 'Above average'))
df$rating <- ordered(df$rating,
                       levels = c('Below average', 'Average', 'Above average'))

head(df,10)

```

```{r echo=FALSE}
str(df)
```

```{r echo=FALSE}
summary(df)

```

#### Our dataset consists of <span style="color:red">13 variables</span> , with almost <span style="color:red">1599 observations</span> 

```{r echo=FALSE}

p1 <- ggplot(df, aes( x = 1, y = pH ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = pH  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)

```

```{r echo=FALSE}
summary(df$pH)
```



```{r echo=FALSE}

p1 <- ggplot(df, aes( x = 1, y = density ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = density  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)

```


#### As we notice in the <span style="color:red">pH and density</span> histogram graphs in red wine quality dataset, the distribution  <span style="color:red">shape is symmetric</span>  and it has the beautiful bell shape which indicate that the data  <span style="color:red">well distributed</span>



```{r echo=FALSE}

p1 <- ggplot(df, aes( x = 1, y = free.sulfur.dioxide ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = free.sulfur.dioxide  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)
```

#### Based on the graph of <span style="color:red"> free.sulfur.dioxide </span>, the most common amount is <span style="color:red"> approximately 7 </span> which. <span style="color:red"> falls in the left side of the distribution </span>


```{r echo=FALSE}

p1 <- ggplot(df, aes( x = 1, y = total.sulfur.dioxide ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = total.sulfur.dioxide  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)

```

```{r echo=FALSE}
summary(df$total.sulfur.dioxide)
```


```{r echo=FALSE}

p1 <- ggplot(df, aes( x = 1, y = sulphates ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = sulphates  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)

```

#### There are <span style="color:red"> more sulphates between 0.0 - 1.0 than 1.1 - 2.0 </span> in the samples available in the dataset.


```{r echo=FALSE}
p1 <- ggplot(df, aes( x = 1, y = alcohol ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = alcohol  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)
```


```{r echo=FALSE}

p1 <- ggplot(df, aes( x = 1, y = fixed.acidity ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = fixed.acidity  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)

```


#### The distribution shape of the <span style="color:red"> free.sulfur.doxide, total.sulfur.doxide, fixed.acidity and alcohol </span> is taking the <span style="color:red"> right skewed shape. </span>



```{r echo=FALSE}

p1 <- ggplot(df, aes( x = 1, y = residual.sugar ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = residual.sugar  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)

```

#### Residual sugar histogram shape is right skewed with most common values between (1-3). However, according to the box polt, Residual sugar has outlier 


```{r echo=FALSE}
p1 <- ggplot(df, aes( x = 1, y = volatile.acidity ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = volatile.acidity  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)


```



```{r echo=FALSE}

p1 <- ggplot(df, aes( x = 1, y = citric.acid ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = citric.acid  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)

```

```{r echo=FALSE}

p1 <- ggplot(df, aes( x = 1, y = chlorides ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) 
  
p2 <- ggplot(df, aes( x   = chlorides  ) ) + 
                   geom_histogram(bins=30 )
  
grid.arrange( p1, p2, ncol=2)

```
#### Chloride histogram shape is right skewed with most common values between (0.1-0.17). However, according to the box polt, Chloride has outlier 


```{r echo=FALSE}
ggplot(data= df, aes(x= quality))+
  geom_bar()
```

#### It seems that the <span style="color:red"> most common quality </span> of the red wine samples will  <span style="color:red"> fall from 5-6 </span> where I would assume that 8 is the best quality based on the physicochemical tests.

#### The interesting question that could be ask here is, <span style="color:red"> what characteristics would be associated with the most common red wine quality? </span> 


# Univariate Analysis


### What is the structure of your dataset?
####There are <span style="color:red">1599 observation </span>  in red wine quality dataset with <span style="color:red">13 attributes </span>  ( fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, pH, sulphates, alcohol, Output variable, quality)

### What is/are the main feature(s) of interest in your dataset?
#### The main features in the dataset is the <span style="color:red"> quality </span>. I would like to know what attributes associated with different wine quality samples. <span style="color:red"> Does the sugar or chloride affect the quality ? </span> I would assume that sugar will definitely affect the quality of the red wine  

### What other features in the dataset do you think will help support your \
####<span style="color:red"> Chlorides , sugar , acidity, Alcohol percentage and density </span> 

### Did you create any new variables from existing variables in the dataset?
#### I created new column and call it <span style="color:red">‘Rating’</span>. Rating column depends entirely on the values entered in the quality column as the following:
#### -	<span style="color:red"> Quality < 5 = ‘Below average’ </span>
#### -	<span style="color:red"> 5 < Quality < 7 = ‘Average’ </span>
#### -	<span style="color:red"> 7 < Quality < 10 = ‘Above average’</span>
 
####The idea behind creating the rating column group all values in 3 categories and start analyzing the data based on the three categories



### Of the features you investigated, were there any unusual distributions? \
####citric.acid distribution has <span style="color:red"> unusual histogram  </span>


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}

cor.plot(select (df,-c(quality, rating)), main = 'Correlation of Red wine dataset') #just blue implies positive manifold

```

#### The fixed acid of the red wine <span style="color:red"> tend to moderately correlate pH, density and critic.acid </span>. 

#### lets now try to subset the data and look to correlation of the worst (3) and best (8) red wine quality in the dataset .


```{r echo=FALSE}

cor.plot(bestWineDf, main = 'Correlation of the best quality red wine') #just blue implies positive manifold

```

### The above table is the subset of the red wine dataset which has on only the data that has the maximum quality (8). We can confirm that density correlate with quality . 

```{r echo=FALSE}

cor.plot(worstWineDf, main = 'Correlation of the worst quality red wine') #just blue implies positive manifold

```

#### In the other hands, the above plot explain the correlation of the red wine worst quality chrematistics. <span style="color:red"> Alcohol percentage have strong correlation with pH </span>.

```{r echo=FALSE}

  ggplot(aes(x = fixed.acidity, 
           y = density   ), 
       data = df) +
      geom_point(alpha = 0.3, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1) 
  
```

#### Density has a very <span style="color:red"> strong correlation with Fixed Acidity. </span>  


```{r echo=FALSE}

  ggplot(aes(x = volatile.acidity, 
           y = pH   ), 
       data = df) +
      geom_point(alpha = 0.3, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1) 
  
```



```{r echo=FALSE}

ggplot(data=df, aes(x=quality, y=alcohol, group= quality, fill=quality)) +
  geom_jitter( alpha = .7, color= 'grey') +
  geom_boxplot(alpha = .5) 

```

#### We can see from the boxplot above that higher quality wine has less outlier

```{r echo=FALSE}

ggplot(data=df, aes(x=quality, y=residual.sugar, group= quality, fill=quality)) +
  geom_jitter( alpha = .7, color= 'grey') +
  geom_boxplot(alpha = .5) +
  scale_y_continuous(lim = c(0,8))

```

#### Wouldn't you assume that <span style="color:red"> sugar always make things taste better? </span> Wouldn't surprise you that <span style="color:red"> in the red wine dataset this is not the case ? </span>

```{r echo=FALSE}
ggplot(data=df, aes(x=quality, y=chlorides, group= quality, fill=quality)) +
  geom_jitter( alpha = .7, color= 'grey') +
  geom_boxplot(alpha = .5) +
  scale_y_continuous(lim = c(0,0.2))
```



### Next, I’ll look at how the categorical features vary with sulfate and qulity
```{r echo=FALSE}
ggplot(data=df, aes(x=quality, y=sulphates, group= quality, fill=quality)) +
  geom_jitter( alpha = .7, color= 'grey') +
  geom_boxplot(alpha = .5) 

```

```{r echo=FALSE}

ggplot(data=df, aes(x=quality, y=pH, group= quality, fill=quality)) +
  geom_jitter( alpha = .7, color= 'grey') +
  geom_boxplot(alpha = .5) 

```

```{r echo=FALSE}

ggplot(data=df, aes(x=quality, y=density, group= quality, fill=quality)) +
  geom_jitter( alpha = .7, color= 'grey') +
  geom_boxplot(alpha = .5) 

```

```{r echo=FALSE}
ggplot(data=df, aes(x=quality, y=fixed.acidity, group= quality, fill=quality)) +
  geom_jitter( alpha = .7, color= 'grey') +
  geom_boxplot(alpha = .5) 


```
 
  
  
# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \investigation. How did the feature(s) of interest vary with other features in \the dataset?

#### - <span style="color:red"> Better wine quality is associated with alcohol </span>
#### - It is worth mentioning that Fixed Acidity  seems to have <span style="color:red">pH </span> almost no effect on quality.
#### - Better wine tends to have <span style="color:red"> less pH concentration  </span> 
#### - Interestingly, <span style="color:red"> residual.sugar has no effect on making great red wine taste </span> I assumed that sugar has strong relationship with better wine quality 


### Did you observe any interesting relationships between the other features \ (not the main feature(s) of interest)?

### What was the strongest relationship you found?
#### - <span style="color:red"> Fixed acidity and density have strong correlation in the red wine quality dataset </span> 

# Multivariate Plots Section

#### Since we have discovered that <span style="color:red"> Alcohol has strong correlation with quality </span>. We would like to find out if alcohol is affected with other attributes.

```{r echo=FALSE, Multivariate_Plots}
ggplot(data = df,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) 
```

#### <span style="color:red"> Sulfate and alcohol percentage did not affect each other </span>. 

```{r echo=FALSE}

ggplot(data = df,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) 
```

#### Well, it seems that <span style="color:red"> less volatile.acidity and alcohol percentage could contribute to better wine quality </span> 

```{r echo=FALSE}

ggplot(data = df,
       aes(y = density, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) 
```

#### The plot above is showing clearly that there <span style="color:red"> is strong negative correlation between better quality and density </span>. 


```{r echo=FALSE}
ggplot(df, aes(x = alcohol, y = density, color = residual.sugar)) +
  geom_jitter() 

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \investigation. Were there features that strengthened each other in terms of \looking at your feature(s) of interest?
#### - <span style="color:red"> alcohol increases and the volitile acidity decreases </span> 
#### - <span style="color:red"> There is strong negative correlation between alcohol and density </span> 

### Were there any interesting or surprising interactions between features?
#### <span style="color:red"> No, there were not any surprise </span> 

### OPTIONAL: Did you create any models with your dataset? Discuss the \ strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data= df, aes(x= density))+
  geom_histogram(fill= 'Pink', col="black")+
  xlab('Density (g / cm^3)') +
  ylab('Number of Samples') + 
  ggtitle("Histogram of Density in Red Wine Samples")
```


### Description One

#### <span style="color:red"> Density values in the red wine quality dataset is normally distributed. </span> Most values is between 0.990 - 1.000. The most common value in the density is 0.997. The shape of the density histogram is symmetric

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data=df, aes(x=quality, y=alcohol, group= quality, fill=quality)) +
  geom_jitter( alpha = .7, color= 'grey') +
  geom_boxplot(alpha = .5)+
  xlab('Quality (1-10)') +
  ylab('Alcohol %') + 
  ggtitle("Influence of Alcohol over Quality")
```

### Description Two

#### <span style="color:red"> Red wine quality 5 has alcohol outliers and it shows that quality 8 </span> has strong correlation with alcohol


### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(df, aes(x = alcohol, y = density, color = residual.sugar)) +
  geom_jitter() +
    xlab('Alcohol %') +
  ylab('Density (g / cm^3)') + 
  ggtitle("Density Correlation with Alcohol ")
```

### Description Three

#### In order to make better wine quality according to the dataset, <span style="color:red"> the density amount should be less as the alcohol percentage increases. </span> 

------

# Reflection
#### - The dataset doesn't have <span style="color:red"> enough data to make valid conclusion </span>. The 1599 records in the dataset is not huge enough to make recommendation.

#### - The dataset did not <span style="color:red"> mention from where they have been collected </span>. I believe the result of this analysis can be applicable to specific geographic location but not another.

#### - <span style="color:red"> Assumption is the mother of all mistake! </span> and this is applicable in the data analyst. I assumed that sugar will definitely have great affect on the quality of the wine. However, as the shown in the plots above. Sugar is not a factor to determine the quality of the red wine


#### - installing R packages on Mac is an exhausting process. It took me 3 days to fix an issue I have in Mac environment. It was preventing me from installing package. I even post a question on StackOverFlow website and finally I was able to fix the issue more details on my post below:

#### https://stackoverflow.com/questions/53680954/rstudio-error-unable-to-find-utility-package-not-a-developer-tool-or-in-path
